---
title: "15 Insights from Wyoming School Enrollment Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{15 Insights from Wyoming School Enrollment Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  eval = TRUE
)
```

```{r load-packages}
library(wyschooldata)
library(dplyr)
library(tidyr)
library(ggplot2)

theme_set(theme_minimal(base_size = 14))

# Load pre-computed data bundled with the package.
# This ensures vignettes build reliably in CI without network access.
# Falls back to live fetch if bundled data is unavailable.
enr <- tryCatch(
  readRDS(system.file("extdata", "enr_2000_2007_tidy.rds", package = "wyschooldata")),
  error = function(e) {
    warning("Bundled data not found, fetching live data")
    fetch_enr_multi(2000:2007, use_cache = TRUE)
  }
)
if (is.null(enr) || nrow(enr) == 0) {
  enr <- fetch_enr_multi(2000:2007, use_cache = TRUE)
}

enr_2007 <- tryCatch(
  readRDS(system.file("extdata", "enr_2007_tidy.rds", package = "wyschooldata")),
  error = function(e) {
    warning("Bundled data not found, fetching live data")
    fetch_enr(2007, use_cache = TRUE)
  }
)
if (is.null(enr_2007) || nrow(enr_2007) == 0) {
  enr_2007 <- fetch_enr(2007, use_cache = TRUE)
}
```

This vignette explores Wyoming's public school enrollment data from the PDF era (2000-2007), the years currently available through Wyoming Department of Education downloads. Wyoming, America's least populous state, offers a unique window into how energy economics shape rural education.

---

## 1. Wyoming lost 4,500 students in seven years, then started bouncing back

Statewide enrollment fell from 90,065 in 2000 to a low of 83,705 in 2005 before recovering to 85,578 in 2007 -- a pattern driven by energy sector volatility.

```{r statewide-data}
state_totals <- enr |>
  filter(is_state, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(end_year, n_students) |>
  mutate(change = n_students - lag(n_students),
         pct_change = round(change / lag(n_students) * 100, 1))

stopifnot(nrow(state_totals) > 0)
state_totals
```

```{r statewide-chart}
ggplot(state_totals, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.2, color = "#654321") +
  geom_point(size = 3, color = "#654321") +
  scale_y_continuous(labels = scales::comma, limits = c(0, NA)) +
  labs(
    title = "Wyoming Public School Enrollment (2000-2007)",
    subtitle = "Enrollment dipped 7% mid-decade before recovering",
    x = "School Year (ending)",
    y = "Total Enrollment"
  )
```

For context: Denver Public Schools alone serves more students than all of Wyoming.

---

## 2. Cheyenne and Casper together educate 29% of Wyoming students

Laramie #1 (Cheyenne) and Natrona #1 (Casper) serve nearly a third of the state -- Wyoming's two largest districts in a landscape of 48 others.

```{r top-districts-data}
top_districts <- enr_2007 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  arrange(desc(n_students)) |>
  head(8) |>
  select(district_name, n_students)

stopifnot(nrow(top_districts) > 0)
top_districts
```

```{r top-districts-chart}
top_districts |>
  mutate(district_name = forcats::fct_reorder(district_name, n_students)) |>
  ggplot(aes(x = n_students, y = district_name, fill = district_name)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n_students)), hjust = -0.1) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.15))) +
  scale_fill_viridis_d(option = "plasma", begin = 0.2, end = 0.8) +
  labs(
    title = "Wyoming's Largest School Districts (2007)",
    subtitle = "Laramie #1 (Cheyenne) and Natrona #1 (Casper) dominate",
    x = "Number of Students",
    y = NULL
  )
```

---

## 3. Sublette County exploded 47% as natural gas boomed

Sublette #1 (Pinedale) grew from 639 to 940 students between 2000 and 2007 -- the fastest growth in the state -- fueled by the Jonah and Pinedale Anticline gas fields.

```{r sublette-data}
sublette <- enr |>
  filter(is_district, district_name == "Sublette #1",
         subgroup == "total_enrollment", grade_level == "TOTAL") |>
  select(end_year, n_students) |>
  mutate(pct_change = round((n_students / lag(n_students) - 1) * 100, 1))

stopifnot(nrow(sublette) > 0)
sublette
```

```{r sublette-chart}
ggplot(sublette, aes(x = end_year, y = n_students)) +
  geom_line(linewidth = 1.2, color = "#CC6600") +
  geom_point(size = 3, color = "#CC6600") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Sublette #1 (Pinedale) Enrollment (2000-2007)",
    subtitle = "Natural gas boom drove 47% enrollment growth",
    x = "School Year (ending)",
    y = "Total Enrollment"
  )
```

When the Jonah gas field ramped up production, Pinedale transformed from a quiet ranching town into a boomtown.

---

## 4. Fremont County lost 14% of its students in seven years

Home to the Wind River Reservation, Fremont County's six school districts shed nearly 1,000 students between 2000 and 2007 -- the steepest county-level decline in the state.

```{r fremont-data}
fremont <- enr |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Fremont", district_name)) |>
  group_by(end_year) |>
  summarize(total = sum(n_students, na.rm = TRUE)) |>
  mutate(pct_change = round((total / lag(total) - 1) * 100, 1))

stopifnot(nrow(fremont) > 0)
fremont
```

```{r fremont-chart}
ggplot(fremont, aes(x = end_year, y = total)) +
  geom_line(linewidth = 1.2, color = "#8B4513") +
  geom_point(size = 3, color = "#8B4513") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Fremont County Enrollment (2000-2007)",
    subtitle = "Wind River Reservation region lost nearly 1,000 students",
    x = "School Year (ending)",
    y = "Total Enrollment"
  )
```

---

## 5. Fremont #38 (Wind River Reservation) lost 39% of its students

Among individual districts, Fremont #38 experienced the most severe decline: from 538 students in 2000 to just 328 by 2007.

```{r fremont38-data}
fremont_districts <- enr |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Fremont", district_name)) |>
  select(end_year, district_name, n_students)

stopifnot(nrow(fremont_districts) > 0)
fremont_districts |>
  filter(end_year %in% c(2000, 2007)) |>
  pivot_wider(names_from = end_year, values_from = n_students) |>
  mutate(change = `2007` - `2000`,
         pct_change = round((`2007` / `2000` - 1) * 100, 1)) |>
  arrange(pct_change)
```

```{r fremont38-chart}
fremont_districts |>
  ggplot(aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Fremont County District Enrollment (2000-2007)",
    subtitle = "Reservation districts experienced steepest declines",
    x = "School Year (ending)",
    y = "Enrollment",
    color = "District"
  )
```

---

## 6. Campbell County: Coal kept Gillette stable while the rest of Wyoming shrank

Campbell #1 (Gillette) held nearly steady at ~7,400 students even as the state lost 7%. The Powder River Basin coal economy provided stability -- a pattern that would reverse dramatically in later years as coal declined.

```{r campbell-data}
campbell <- enr |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Campbell", district_name)) |>
  select(end_year, district_name, n_students)

campbell_summary <- campbell |>
  group_by(end_year) |>
  summarize(total = sum(n_students, na.rm = TRUE)) |>
  mutate(pct_change = round((total / lag(total) - 1) * 100, 1))

stopifnot(nrow(campbell_summary) > 0)
campbell_summary
```

```{r campbell-chart}
ggplot(campbell_summary, aes(x = end_year, y = total)) +
  geom_line(linewidth = 1.2, color = "#2F4F4F") +
  geom_point(size = 3, color = "#2F4F4F") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Campbell County (Gillette) Enrollment (2000-2007)",
    subtitle = "Coal economy kept enrollment stable during statewide decline",
    x = "School Year (ending)",
    y = "Total Enrollment"
  )
```

---

## 7. Kindergarten overtook 12th grade: Wyoming's demographic crossover

In 2000, Wyoming graduated 6,851 seniors but enrolled only 5,825 kindergartners. By 2007 the pipeline had flipped: 6,891 kindergartners vs 6,212 12th graders. The state was getting younger at the bottom.

```{r k-12-pipeline}
k_vs_12 <- enr |>
  filter(is_state, subgroup == "total_enrollment", grade_level %in% c("K", "12")) |>
  select(end_year, grade_level, n_students) |>
  pivot_wider(names_from = grade_level, values_from = n_students)

stopifnot(nrow(k_vs_12) > 0)
k_vs_12
```

```{r k-12-chart}
k_vs_12_long <- enr |>
  filter(is_state, subgroup == "total_enrollment", grade_level %in% c("K", "12")) |>
  select(end_year, grade_level, n_students)

ggplot(k_vs_12_long, aes(x = end_year, y = n_students, color = grade_level)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_manual(values = c("K" = "#0072B2", "12" = "#D55E00")) +
  labs(
    title = "Kindergarten vs 12th Grade Enrollment (2000-2007)",
    subtitle = "K enrollment overtook 12th grade by 2005",
    x = "School Year (ending)",
    y = "Enrollment",
    color = "Grade"
  )
```

---

## 8. Wyoming consolidated 28 schools in seven years

From 382 schools in 2000 to 354 in 2007, Wyoming lost 7% of its school buildings. Rural school consolidation was reshaping the education landscape even as enrollment began recovering.

```{r school-count-data}
school_counts <- enr |>
  filter(is_school, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  group_by(end_year) |>
  summarize(n_schools = n(), total_students = sum(n_students, na.rm = TRUE)) |>
  mutate(avg_size = round(total_students / n_schools))

stopifnot(nrow(school_counts) > 0)
school_counts
```

```{r school-count-chart}
ggplot(school_counts, aes(x = end_year, y = n_schools)) +
  geom_line(linewidth = 1.2, color = "#009E73") +
  geom_point(size = 3, color = "#009E73") +
  scale_y_continuous(limits = c(0, NA)) +
  labs(
    title = "Number of Wyoming Schools (2000-2007)",
    subtitle = "28 schools closed or consolidated in seven years",
    x = "School Year (ending)",
    y = "Number of Schools"
  )
```

---

## 9. Sweetwater County's twin cities: Rock Springs held, Green River shrank

Rock Springs (Sweetwater #1) and Green River (Sweetwater #2) are 15 miles apart but had different enrollment trajectories. Rock Springs recovered by 2007; Green River kept declining.

```{r sweetwater-data}
sweetwater <- enr |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Sweetwater", district_name)) |>
  select(end_year, district_name, n_students)

stopifnot(nrow(sweetwater) > 0)
sweetwater
```

```{r sweetwater-chart}
sweetwater |>
  ggplot(aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Sweetwater County Districts (2000-2007)",
    subtitle = "Rock Springs recovered; Green River kept declining",
    x = "School Year (ending)",
    y = "Enrollment",
    color = "District"
  )
```

---

## 10. Teton County (Jackson Hole) stayed flat while the rest of Wyoming shrank

In a state where most districts were losing students, Teton #1 held steady around 2,200-2,400 -- Jackson's tourism and recreation economy insulated it from the broader decline.

```{r teton-data}
teton <- enr |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL",
         grepl("Teton", district_name)) |>
  select(end_year, district_name, n_students)

stopifnot(nrow(teton) > 0)
teton
```

```{r teton-chart}
teton |>
  ggplot(aes(x = end_year, y = n_students, color = district_name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Teton County (Jackson) Enrollment (2000-2007)",
    subtitle = "Tourism economy kept enrollment stable",
    x = "School Year (ending)",
    y = "Enrollment",
    color = "District"
  )
```

---

## 11. Elementary enrollment grew while high school shrank

Between 2000 and 2007, elementary (K-5) enrollment rose from 38,545 to 39,336 (+2%) while high school (9-12) dropped from 30,172 to 26,839 (-11%). The pipeline was filling from the bottom.

```{r elem-hs-data}
elem <- enr |>
  filter(is_state, subgroup == "total_enrollment",
         grade_level %in% c("K", "01", "02", "03", "04", "05")) |>
  group_by(end_year) |>
  summarize(students = sum(n_students)) |>
  mutate(level = "Elementary (K-5)")

hs <- enr |>
  filter(is_state, subgroup == "total_enrollment",
         grade_level %in% c("09", "10", "11", "12")) |>
  group_by(end_year) |>
  summarize(students = sum(n_students)) |>
  mutate(level = "High School (9-12)")

elem_hs <- bind_rows(elem, hs)

stopifnot(nrow(elem_hs) > 0)
elem_hs
```

```{r elem-hs-chart}
ggplot(elem_hs, aes(x = end_year, y = students, color = level)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_manual(values = c("Elementary (K-5)" = "#0072B2", "High School (9-12)" = "#D55E00")) +
  labs(
    title = "Elementary vs High School Enrollment (2000-2007)",
    subtitle = "Elementary growing while high school shrank",
    x = "School Year (ending)",
    y = "Enrollment",
    color = "Level"
  )
```

---

## 12. Vast distances define Wyoming education: 48 districts across 97,000 square miles

Wyoming's 48 school districts average about 1,780 students each, but cover an average of 2,000 square miles per district. Many districts are geographically larger than some eastern states.

```{r district-size}
n_districts <- enr_2007 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  summarize(
    n_districts = n(),
    total_students = sum(n_students, na.rm = TRUE),
    avg_per_district = round(total_students / n_districts)
  )

stopifnot(nrow(n_districts) > 0)
n_districts
```

With an average of fewer than 1,800 students per district, Wyoming's districts are intimate by national standards but geographically immense.

---

## 13. Small schools are the Wyoming norm: 109 schools have under 100 students

In 2007, 109 out of 354 schools (31%) served fewer than 100 students. Wyoming pays to keep tiny schools open across the frontier rather than bus children hours to a larger campus.

```{r small-schools}
school_sizes <- enr_2007 |>
  filter(is_school, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  mutate(size_category = case_when(
    n_students < 50 ~ "Under 50",
    n_students < 100 ~ "50-99",
    n_students < 250 ~ "100-249",
    n_students < 500 ~ "250-499",
    TRUE ~ "500+"
  )) |>
  count(size_category, name = "n_schools") |>
  mutate(size_category = factor(size_category,
         levels = c("Under 50", "50-99", "100-249", "250-499", "500+")))

stopifnot(nrow(school_sizes) > 0)
school_sizes
```

```{r small-schools-chart}
ggplot(school_sizes, aes(x = size_category, y = n_schools, fill = size_category)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = n_schools), vjust = -0.3) +
  scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.8) +
  labs(
    title = "Wyoming School Size Distribution (2007)",
    subtitle = "109 schools serve fewer than 100 students",
    x = "School Size (students)",
    y = "Number of Schools"
  )
```

---

## 14. Grade-level enrollment: 9th grade bulge reflects the demographic pipeline

The 2007 grade-level snapshot shows a pronounced 9th grade bulge (7,069 students) that shrinks to 6,212 by 12th grade -- a pattern suggesting either dropouts or outmigration in the upper grades.

```{r grade-level-data}
grade_enr <- enr_2007 |>
  filter(is_state, subgroup == "total_enrollment",
         grade_level %in% c("K", "01", "02", "03", "04", "05",
                            "06", "07", "08", "09", "10", "11", "12")) |>
  select(grade_level, n_students) |>
  mutate(grade_level = factor(grade_level,
         levels = c("K", "01", "02", "03", "04", "05",
                    "06", "07", "08", "09", "10", "11", "12")))

stopifnot(nrow(grade_enr) > 0)
grade_enr
```

```{r grade-level-chart}
ggplot(grade_enr, aes(x = grade_level, y = n_students, fill = grade_level)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n_students)), vjust = -0.3, size = 3) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.1))) +
  scale_fill_viridis_d(option = "viridis") +
  labs(
    title = "Wyoming Enrollment by Grade Level (2007)",
    subtitle = "9th grade bulge shrinks by 12% through senior year",
    x = "Grade Level",
    y = "Number of Students"
  )
```

---

## 15. The smallest districts: Washakie #2 has just 96 students

Wyoming keeps schools open in communities so small that the entire district could fit in a single classroom. Washakie #2 (Ten Sleep) enrolled 96 students in 2007; Sheridan #3 (Clearmont) had 101.

```{r smallest-districts-data}
smallest <- enr_2007 |>
  filter(is_district, subgroup == "total_enrollment", grade_level == "TOTAL") |>
  arrange(n_students) |>
  head(10) |>
  select(district_name, n_students)

stopifnot(nrow(smallest) > 0)
smallest
```

```{r smallest-districts-chart}
smallest |>
  mutate(district_name = forcats::fct_reorder(district_name, n_students)) |>
  ggplot(aes(x = n_students, y = district_name, fill = district_name)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = scales::comma(n_students)), hjust = -0.1) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.2))) +
  scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.8) +
  labs(
    title = "Wyoming's Smallest School Districts (2007)",
    subtitle = "One-school towns keeping education local in the rural West",
    x = "Number of Students",
    y = NULL
  )
```

---

## Summary

Wyoming's 2000-2007 school enrollment data reveals:

- **America's smallest system**: Under 90,000 students statewide
- **Energy-driven volatility**: State enrollment fell 7% before recovering with the gas boom
- **Two-city dominance**: Cheyenne and Casper together serve 29% of students
- **Gas boom town**: Sublette County (Pinedale) grew 47% as natural gas production surged
- **Reservation decline**: Fremont County lost 14% of students, with some districts losing nearly 40%
- **Coal stability**: Campbell County held steady in an era before coal's decline
- **Demographic crossover**: Kindergarten enrollment overtook 12th grade by 2005
- **School consolidation**: 28 schools closed between 2000 and 2007
- **Small school tradition**: 109 of 354 schools serve under 100 students
- **Tiny districts**: Washakie #2 (Ten Sleep) enrolled just 96 students

These patterns reflect Wyoming's unique position as America's least populous state, where energy economics and vast distances shape every aspect of public education.

---

*Data sourced from the Wyoming Department of Education [edu.wyoming.gov](https://edu.wyoming.gov/). Currently available years: 2000-2007 (PDF era). Modern era data (2008+) from the WDE reporting portal is not currently accessible.*

---

## Session Info

```{r session-info}
sessionInfo()
```
